Learning a functional distance between organizations with unsupervised embeddings of scientific mobility

### 2020-08-03 paper/current/main.tex
implenet sada comments—data and methods

%% word2vec
These affiliation trajectories are used as input to the standard skip-gram negative sampling word embedding, commonly known as \textit{word2vec}~\autocite{mikolov2013word2vec}.
\textit{word2vec} constructs dense and continuous vector representations of words and phrases, in which distance between words corresponds to a notion of semantic distance. By embedding affiliation trajectories, we aim to learn a dense vector for every organization where the distance between vectors relates to the tendency for two organizations to occur in similar contexts.
Formally,  the model attempts to maximize average log probability of a target organization appearing in an affiliation trajectory $a_{t}$, given organizations in the surrounding context window, $a_{t+k}$. This can be written as follow, where $w$ is the size of the window and $T$ iterates through possible target organizations:

%
% MAIN W2V Equation
%
$$
\frac{1}{T}\sum_{t = 1}^{T} \sum_{-w \leq j \leq w, j \neq 0} \log p(a_{t + j} \vert a_{t}),
$$
where the conditional probability of observing one organization $a_{i}$, given another $a_{j}$, is defined as,

%
% CONDITIONAL PROBABILITY EQ
%
$$
p(a_{i} \vert a_{j}) = \frac{\exp(\bm{u}_{a_{i}} \cdot \bm{v}_{a_{j}})}{\sum_{n=1}^N \exp(\bm{u}_{a_{n}} \cdot \bm{v}_{a_{j}})},
$$
where $N$ is the total number of organizations and $\bm{v}$ and $\bm{u}$ are the ``in-vector" and ``out-vector" respectively~\autocite{mikolov2013word2vec}.  We follow past studies and use the in-vector, $\bm{v}$, as the final representation~\autocite{linzhuo2020hyperbolic, tshitoyan2019mat2vec, garg2018gender, kozlowski2018geometry, hamilton2016diachronic, le2014doc2vec, nakandala2016twitch}.

%% Word2vec parameters and training procedure
We used the \texttt{gensim} python implementation of the \textit{word2vec} model. 
The skip-gram negative sampling  \textit{word2vec} model has several tunable hyper-parameters, including the embedding dimension $d$, the size of the context window $w$, the minimum frequency threshold $f_{\min}$, initial learning rate $\alpha$, and the number of iterations.
For main results, we used $d=300$ and $w=1$, which were the parameters that best explained the flux between organizations, though results were robust across different settings (Fig.~\ref{fig:supp:hyperparams}).
To mitigate the effect of less common organizations, we set the $f_{\min}$ as 50, limiting to organizations appearing at least 50 times across the training trajectories;
6,580 unique organizations appear in the resulting embedding.
We set $\alpha$ to its default value of 0.025 and iterate five times over all training affiliation trajectories.

>>>>

%% word2vec
These affiliation trajectories are used as input to the standard skip-gram negative sampling word embedding, commonly known as \textit{word2vec}~\autocite{mikolov2013word2vec}.
\textit{word2vec} constructs dense and continuous vector representations of words and phrases, in which distance between words corresponds to a notion of semantic distance. By embedding affiliation trajectories, we aim to learn a dense vector for every organization where the distance between vectors relates to the tendency for two organizations to occur in similar contexts.
Suppose an affiliation trajectory, denoted by ($a_{1}, a_{2}, \ldots, a_{T}$), where $a_{t}$ is the $t$th organization in the trajectory. An organization, $a_{t}$, is considered to have context organizations, $a_{t-w}, \ldots, a_{t-1}, a_{t+1},\ldots, a_{t+w}$, that appear in the window surrounding $a_t$ up to a time lag of $w$, where $w$ is the window size parameter truncated at $t - w \geq 0$ and $t + w \leq T$. Then, the model learns probability $p(a_{t + \tau} \vert a_{t})$, where $-w\leq \tau\leq w$ and $\tau \neq 0$,  by maximizing its log likelihood given by

%
% MAIN W2V Equation
%
\begin{equation}
\frac{1}{T}\sum_{t = 1}^{T} \sum_{-w \leq \tau \leq w, \tau \neq 0} \log p(a_{t + \tau} \vert a_{t}),
\end{equation}
where,

%
% CONDITIONAL PROBABILITY EQ
%
\begin{equation}
p(a_{i} \vert a_{j}) = \frac{\exp(\bm{u}_{i} \cdot \bm{v}_{a_{j}})}{\sum_{n=1}^N \exp(\bm{u}_{n} \cdot \bm{v}_{a_{j}})},
\end{equation}
where $a_{t+\tau} = j \vert a_{t+\tau} = i$, and $N$ is the number of unique organizations represented in the data, and $\bm{v}$ and $\bm{u}$ are the ``in-vector" and ``out-vector" respectively.
The in-vector and out-vectors are different kinds of vector representations for organizations. We only use the in-vector, $v$, because the in-vector is more widely used and often superiors to the out-vector in link prediction benchmarks~\autocite{linzhuo2020hyperbolic, tshitoyan2019mat2vec, garg2018gender, kozlowski2018geometry, hamilton2016diachronic, le2014doc2vec, nakandala2016twitch}.

%% Word2vec parameters and training procedure
We used the \textit{word2vec}  implenetation in the python package \texttt{gensim}. 
The skip-gram negative sampling  \textit{word2vec} model has several tunable hyper-parameters, including the embedding dimension $d$, the size of the context window $w$, the minimum frequency threshold $f_{\min}$, initial learning rate $\alpha$, and the number of iterations.
For main results, we used $d=300$ and $w=1$, which were the parameters that best explained the flux between organizations, though results were robust across different settings (Fig.~\ref{fig:supp:hyperparams}).
To mitigate the effect of less common organizations, we set $f_{\min} = 50$, limiting to organizations appearing at least 50 times across the training trajectories;
6,580 unique organizations appear in the resulting embedding.
We set $\alpha$ to its default value of 0.025 and iterate five times over all training affiliation trajectories.
